NumLin: Linear Types for Linear Algebra

### 2019-01-09 write-up/paper/discussion.tex
Beefed up the related work

Using linear types for BLAS routines is a particularly good domain fit (given
the implicit restrictions on aliasing arguments), but is under-explored in the
academic literature. Linear algebra libraries written in Rust take advantage of
the distinction that Rust's type system offers between mutable views/references
to arrays, offering a \emph{seemingly} different approach to the
fractional-permissions based one we took in this paper.  However, recent
research\cite{weiss} suggests that Rust's borrow-checker could be explained in
simpler terms using \emph{fractional-permissions}; in this light, our paper
neatly bridges the gap by demonstrating the theoretical and practical
simplicity of this way of thinking.

Linear Haskell\cite{bernardy2017linear}, takes a slightly different definition
of linearity, that is one on \emph{arrows} as opposed to \emph{kinds}: for $f :
a \multimap b$, if $f u$ is used exactly once \emph{then} $u$ is used exactly
once. Whilst this has the advantage of being backwards-compatible, \lang's
approach of compile-time code-generation enables using linearity \emph{as a
library}, whilst the use of fractions gives extra control over aliasing not
available in Linear Haskell.

In general, using substructural types to express array computations is not
particularly new\cite{scholz,henriksen,bernardy2016}, but these approaches
limit recursion and higher-order functions, and have special constructs to
enable \emph{implementation} of efficient BLAS-like libraries rather than
enforcing correct \emph{usage}.

>>>>

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions facilitating compilation to GPUs. The work of
\citet{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

In contrast, our approach is to begin with the assumption that we can
take existing efficient BLAS-like libraries, and then enforce their
correct \emph{usage} using a linear type discipline with fractional
permissions.

This approach is similar to the one taken in linear algebra libraries
for Rust -- these libraries typically take advantage of the
distinction that Rust's type system offers between mutable
views/references to arrays.  The work of \citet{weiss} and
\citet{rustbelt} suggest that Rust's borrow-checker can be expressed
in simpler terms using \emph{fractional-permissions}, though to our
knowledge the programmer-visible lifetime analysis in Rust has never
been formalized.

Using fractional permissions directly, as we did, rather than relying
on a complex has two main benefits. First, our type system
demonstrates that type systems for fractional permissions can be
\emph{dramatically} simpler than existing state-of-the-art approaches,
including both industrial languages like Rust, as well as the academic
languages developed by \citet{bierhoff}.  Their type system, much like
Rust, build a complex dataflow analysis into the typing rules to infer
when variables can be shared or not. This allows for very
natural-looking user programs, but creates the impression that using
fractional permissions requires a heavy theoretical and engineering
effort going well beyond that needed for supporting basic linear
types.

Instead, our approach of requiring sharing to be made explicit means
that the existing unification machinery already in place for ordinary
ML-style type inference, can be reused to support fractions. Basically,
we can view sharing a value as dividing a fraction by two, and after
taking logarithms all fractions are Peano numbers, whose equality can
be established with ordinary unification.

This fact is important because there are major upcoming
implementations of linear types such as Linear
Haskell~\cite{bernardy2017linear}, which do not have built-in support
for fractional permissions. Instead, Linear Haskell takes a slightly
different definition of linearity, one based on \emph{arrows} as
opposed to \emph{kinds}: for $f : a \multimap b$, if $f u$ is used
exactly once \emph{then} $u$ is used exactly once. Whilst this has the
advantage of being backwards-compatible, it also means that the type
system has no built-in support for the concurrent reader, exclusive
writer pattern that fractional permissions enable.

However, since our type system shows demonstrates unification is ``all
one needs'' for fractions, it should be possible to \emph{encode}
\lang's approach to fractional permissions in Linear Haskell by adding
a GADT-style natural number index to array types tracking the
fraction, which should enable supporting high-performance BLAS
bindings in Linear Haskell. Actually implementing this is something we
leave for future work, as there remains an issue which we do not see a
good encoding for. Namely, only having support for linear functions
makes it a bit inconvenient to manipulate linear values directly --
programs end up taking on a CPS-like structure. This seems to remain
an advantage of a direct implementation of linear types.
